---
title: "4plex_Run1"
author: "Jeel Patel"
date: "24.02.2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}

# Install packages if not already installed
if (!requireNamespace("Seurat", quietly = TRUE)) install.packages("Seurat")
if (!requireNamespace("qs", quietly = TRUE)) install.packages("qs")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

# Load libraries
library(Seurat)
library(qs)
library(dplyr)

```
```



```{r define_sample_paths}

# Define paths to the filtered_feature_bc_matrix directories for Run1
sample_paths <- list(
  "S1_W1_R1" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R1_count/outs/filtered_feature_bc_matrix",
  "S1_W1_R2" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R2_count/outs/filtered_feature_bc_matrix",
  "S1_W1_R3" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R3_count/outs/filtered_feature_bc_matrix",
  "S1_W1_R4" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R4_count/outs/filtered_feature_bc_matrix"
)

# Print the sample paths to confirm
print(sample_paths)

```

```
```{r load_samples}
# Define RUN flag (set to TRUE for first run, FALSE for subsequent runs)
RUN <- FALSE

# Specify the results directory
results_dir <- "PBMC_Run1_4plex_results"

# Ensure the results directory exists
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Initialize list to store Seurat objects
seurat_objects <- list()

# Loop through sample paths
for (sample_name in names(sample_paths)) {
  
  # Define file path for the saved .qs file
  qs_file <- paste0(results_dir, "/", sample_name, "_raw.qs")
  
  if (RUN) {
    # Read 10X data
    data_dir <- sample_paths[[sample_name]]
    sample_data <- Read10X(data.dir = data_dir)
    
    # Create Seurat object
    seurat_obj <- CreateSeuratObject(counts = sample_data, 
                                     project = paste0("PBMC_Run1_", sample_name), 
                                     min.cells = 3, 
                                     min.features = 200)
    
    # Save Seurat object using qs::qsave
    qs::qsave(seurat_obj, file = qs_file)
    print(paste("Processed and saved:", sample_name))
    
  } else {
    # Load pre-saved Seurat object using qs::qread
    seurat_obj <- qs::qread(qs_file)
    print(paste("Loaded from file:", sample_name))
  }
  
  # Store Seurat object in list
  seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm loaded objects
print(names(seurat_objects))

```

```{r Seurat_obj}
# Print Seurat Object Summary
seurat_obj

# View the structure
str(seurat_obj)

# View metadata for S1_W1_R1
head(seurat_objects[["S1_W1_R1"]]@meta.data)

# View metadata for S1_W1_R2
head(seurat_objects[["S1_W1_R2"]]@meta.data)
```



################################################################################
##################### ### Without saving PNG images ############################


```{r qc_step_filtered}


### Without saving PNG images

library(Seurat)


# Define QC RUN flag (set to TRUE for first run, FALSE for subsequent runs)
QC_RUN <- TRUE

# Initialize list to store filtered Seurat objects
seurat_objects_filtered <- list()

# Loop through each Seurat object for QC
for (sample_name in names(seurat_objects)) {
  
  # Define file path for post-QC .qs file
  qc_qs_file <- paste0(results_dir, "/", sample_name, "_postQC.qs")
  
  if (QC_RUN) {
    # Load raw Seurat object
    seurat_obj_raw <- seurat_objects[[sample_name]]
    
    # Calculate percent mitochondrial genes
    seurat_obj_raw[["percent.mt"]] <- PercentageFeatureSet(seurat_obj_raw, pattern = "^MT-")
    
    # Load required libraries
library(ggplot2)
library(patchwork)
    
    # Visualize QC metrics (on raw data)
    print(VlnPlot(seurat_obj_raw, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3))

   # Scatter plots for relationships — Wrapped in print()
    p1 <- FeatureScatter(seurat_obj_raw, feature1 = "nCount_RNA", feature2 = "percent.mt")
    p2 <- FeatureScatter(seurat_obj_raw, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
    print(p1 + p2)
    
    # Apply QC filters to create a filtered object (adjust thresholds as needed)
    seurat_obj_filtered <- subset(seurat_obj_raw, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
    
    # Save post-QC Seurat object
    qs::qsave(seurat_obj_filtered, qc_qs_file)
    print(paste("QC completed and saved for:", sample_name))
    
  } else {
    # Load pre-saved post-QC Seurat object
    seurat_obj_filtered <- qs::qread(qc_qs_file)
    print(paste("Loaded post-QC object for:", sample_name))
  }
  
  # Store filtered object in a separate list
  seurat_objects_filtered[[sample_name]] <- seurat_obj_filtered
}

# Confirm QC completion
print("QC step completed for all samples.")

```


####################### Just to visualize Violin plot ##########################

```{r plot_external, echo=TRUE, fig.show='hide'}
VlnPlot(seurat_obj_raw, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

```




################################################################################
######################### With saving PNG images ###############################


```{r qc_step_filtered}

### With saving PNG images

# Define QC RUN flag (set to TRUE for first run, FALSE for subsequent runs)
QC_RUN <- TRUE

# Load required libraries
library(ggplot2)
library(patchwork)
library(Seurat)

# Initialize list to store filtered Seurat objects
seurat_objects_filtered <- list()

# Loop through each Seurat object for QC
for (sample_name in names(seurat_objects)) {
  
  # Define file paths
  qc_qs_file <- paste0(results_dir, "/", sample_name, "_postQC.qs")
  vln_plot_file <- paste0(results_dir, "/Visualize_QC_metrics/", sample_name, "_vln_plot.png")
scatter_plot_file <- paste0(results_dir, "/Visualize_QC_metrics/", sample_name, "_scatter_plot.png")

  
  if (QC_RUN) {
    # Load raw Seurat object
    seurat_obj_raw <- seurat_objects[[sample_name]]
    
    # Calculate percent mitochondrial genes
    seurat_obj_raw[["percent.mt"]] <- PercentageFeatureSet(seurat_obj_raw, pattern = "^MT-")
    
    ## ---- SAVE VIOLIN PLOT ---- ##
    vln_plot <- VlnPlot(seurat_obj_raw, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
    ggsave(filename = vln_plot_file, plot = vln_plot, width = 10, height = 5, dpi = 300)
    print(paste("Violin plot saved for:", sample_name))
    
    ## ---- SAVE SCATTER PLOT ---- ##
    p1 <- FeatureScatter(seurat_obj_raw, feature1 = "nCount_RNA", feature2 = "percent.mt")
    p2 <- FeatureScatter(seurat_obj_raw, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
    scatter_plot <- p1 + p2
    ggsave(filename = scatter_plot_file, plot = scatter_plot, width = 10, height = 5, dpi = 300)
    print(paste("Scatter plot saved for:", sample_name))
    
    # Apply QC filters to create a filtered object
    seurat_obj_filtered <- subset(seurat_obj_raw, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
    
    # Save post-QC Seurat object
    qs::qsave(seurat_obj_filtered, qc_qs_file)
    print(paste("QC completed and saved for:", sample_name))
    
  } else {
    # Load pre-saved post-QC Seurat object
    seurat_obj_filtered <- qs::qread(qc_qs_file)
    print(paste("Loaded post-QC object for:", sample_name))
    
    ## ---- DISPLAY SAVED PLOTS ---- ##
    # Violin Plot
    grid::grid.raster(png::readPNG(vln_plot_file))
    # Scatter Plot
    grid::grid.raster(png::readPNG(scatter_plot_file))
  }
  
  # Store filtered object in a separate list
  seurat_objects_filtered[[sample_name]] <- seurat_obj_filtered
}

# Confirm QC completion
print("QC step completed for all samples.")


```



```{r Normalized_data}


# Define Normalization RUN flag (set to TRUE for first run, FALSE to load saved objects)
NORM_RUN <- TRUE

# Initialize list to store normalized Seurat objects
seurat_objects_normalized <- list()

# Loop through each post-QC Seurat object for normalization
for (sample_name in names(seurat_objects_filtered)) {
  
  # Define file path for post-normalization .qs file
  norm_qs_file <- paste0(results_dir, "/", sample_name, "_normalized.qs")
  
  if (NORM_RUN) {
    # Load post-QC Seurat object
    seurat_obj_filtered <- seurat_objects_filtered[[sample_name]]
    
    # Apply LogNormalization
    seurat_obj_norm <- NormalizeData(seurat_obj_filtered, normalization.method = "LogNormalize", scale.factor = 10000)
    
    # Save normalized Seurat object using qs::qsave
    qs::qsave(seurat_obj_norm, norm_qs_file)
    print(paste("Normalization completed and saved for:", sample_name))
    
  } else {
    # Load pre-saved normalized Seurat object
    seurat_obj_norm <- qs::qread(norm_qs_file)
    print(paste("Loaded normalized object for:", sample_name))
  }
  
  # Store normalized object in list
  seurat_objects_normalized[[sample_name]] <- seurat_obj_norm
}

# Confirm normalization completion
print("Normalization step completed for all samples.")


```




```{r str_seurat_obj_norm}

str(seurat_obj_norm)

```



```{r feature_selection}

# Define Feature Selection RUN flag (set to TRUE for first run, FALSE to load saved objects)
FEATURE_RUN <- TRUE

# Initialize list to store Seurat objects with variable features
seurat_objects_variable <- list()

# Loop through each normalized Seurat object for feature selection
for (sample_name in names(seurat_objects_normalized)) {
  
  # Define file path for post-feature selection .qs file
  feature_qs_file <- paste0(results_dir, "/", sample_name, "_variable_features.qs")
  
  if (FEATURE_RUN) {
    # Load normalized Seurat object
    seurat_obj_norm <- seurat_objects_normalized[[sample_name]]
    
    # Identify highly variable features
    seurat_obj_var <- FindVariableFeatures(seurat_obj_norm, selection.method = "vst", nfeatures = 2000)
    
    # Save Seurat object with variable features
    qs::qsave(seurat_obj_var, feature_qs_file)
    print(paste("Feature selection completed and saved for:", sample_name))
    
    # Identify the top 10 most highly variable genes
    top10 <- head(VariableFeatures(seurat_obj_var), 10)
    
    # Visualize variable features
    plot1 <- VariableFeaturePlot(seurat_obj_var)
    labeled_plot <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
    
    
    # Save the plot as PNG
    plot_file <- paste0(results_dir, "/Visualize_Feature_Selection/", sample_name, "_variable_features.png")
    ggsave(plot_file, labeled_plot, width = 8, height = 6)
    print(paste("Variable feature plot saved for:", sample_name))
    
  } else {
    # Load pre-saved Seurat object with variable features
    seurat_obj_var <- qs::qread(feature_qs_file)
    print(paste("Loaded Seurat object with variable features for:", sample_name))
  }
  
  # Store Seurat object in the list
  seurat_objects_variable[[sample_name]] <- seurat_obj_var
}

# Confirm feature selection completion
print("Feature selection step completed for all samples.")

```



