---
title: "PBMC_Run1_4plex"
author: "Jeel Patel"
date: "2025-02-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r load_libraries}
# Install packages if not already installed
if (!requireNamespace("Seurat", quietly = TRUE)) install.packages("Seurat")
if (!requireNamespace("qs", quietly = TRUE)) install.packages("qs")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")

# Load libraries
library(Seurat)
library(qs)
library(dplyr)

```


######################### Define Sample Paths ##################################

```{r define_sample_paths}

# Define paths to the filtered_feature_bc_matrix directories for Run1
sample_paths <- list(
  "S1_W1_R1" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R1_count/outs/filtered_feature_bc_matrix",
  "S1_W1_R2" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R2_count/outs/filtered_feature_bc_matrix",
  "S1_W1_R3" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R3_count/outs/filtered_feature_bc_matrix",
  "S1_W1_R4" = "/tank2/jpatel/CR_count_results/Individual_Samples/4plex_subpool1_run1_CRC/S1_W1_R4_count/outs/filtered_feature_bc_matrix"
)

# Print the sample paths to confirm
print(sample_paths)

```


#################### Load sample and create seurat onject ######################


```{r load_samples}
# Define RUN flag (set to TRUE for first run, FALSE for subsequent runs)
RUN <- FALSE

# Specify the results directory
results_dir <- "PBMC_Run1_4plex_results"

# Ensure the results directory exists
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Initialize list to store Seurat objects
seurat_objects <- list()

# Loop through sample paths
for (sample_name in names(sample_paths)) {
  
  # Define file path for the saved .qs file
  qs_file <- paste0(results_dir, "/", sample_name, "_raw.qs")
  
  if (RUN) {
    # Read 10X data
    data_dir <- sample_paths[[sample_name]]
    sample_data <- Read10X(data.dir = data_dir)
    
    # Create Seurat object
    seurat_obj <- CreateSeuratObject(counts = sample_data, 
                                     project = paste0("PBMC_Run1_", sample_name), 
                                     min.cells = 3, 
                                     min.features = 200)
    
    # Save Seurat object using qs::qsave
    qs::qsave(seurat_obj, file = qs_file)
    print(paste("Processed and saved:", sample_name))
    
  } else {
    # Load pre-saved Seurat object using qs::qread
    seurat_obj <- qs::qread(qs_file)
    print(paste("Loaded from file:", sample_name))
  }
  
  # Store Seurat object in list
  seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm loaded objects
print(names(seurat_objects))

```

########################## structure of seurat object ##########################

```{r Explore_structure}

# Print Seurat Object Summary
seurat_obj

# View the structure
str(seurat_obj)

# View metadata for S1_W1_R1
head(seurat_objects[["S1_W1_R1"]]@meta.data)

# View metadata for S1_W1_R2
head(seurat_objects[["S1_W1_R2"]]@meta.data)

# View metadata for S1_W1_R3
head(seurat_objects[["S1_W1_R3"]]@meta.data)

# View metadata for S1_W1_R4
head(seurat_objects[["S1_W1_R4"]]@meta.data)

```




```{r}

# View metadata for samples
nrow(seurat_objects[["S1_W1_R1"]]@meta.data)
nrow(seurat_objects[["S1_W1_R2"]]@meta.data)
nrow(seurat_objects[["S1_W1_R3"]]@meta.data)
nrow(seurat_objects[["S1_W1_R4"]]@meta.data)


```


####################### Add percent.mt -> seurat_objects #######################



```{r check_MT_before_QC, fig.width=10, fig.height=5}


# Loop through each Seurat object for QC
for (sample_name in names(seurat_objects)) {
  
 
    # Load raw Seurat object
    seurat_obj <- seurat_objects[[sample_name]]
    
    # Calculate percent mitochondrial genes
    seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
    
     ## ---- SAVE VIOLIN PLOT ---- ##
    vln_plot <- VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
    print(vln_plot)
    
    ## ---- SAVE SCATTER PLOT ---- ##
    p1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
    p2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
    scatter_plot <- p1 + p2
    print(scatter_plot)
    
    # Store filtered object in the list
    seurat_objects[[sample_name]] <- seurat_obj
 
}   
    
```

```{r percent.mt_added}

head(seurat_objects[["S1_W1_R1"]]@meta.data, 5)
head(seurat_objects[["S1_W1_R2"]]@meta.data, 5)
head(seurat_objects[["S1_W1_R3"]]@meta.data, 5)
head(seurat_objects[["S1_W1_R4"]]@meta.data, 5)

```



```{r check_metadata_without_QC}

# View metadata for samples

nrow(seurat_objects[["S1_W1_R1"]]@meta.data)
nrow(seurat_objects[["S1_W1_R2"]]@meta.data)
nrow(seurat_objects[["S1_W1_R3"]]@meta.data)
nrow(seurat_objects[["S1_W1_R4"]]@meta.data)

```

######################### Quality control filters ##############################


```{r post_QC, fig.width=10, fig.height=5}

# Loop through each Seurat object for QC
for (sample_name in names(seurat_objects)) {
  
 
    # Load raw Seurat object
    seurat_obj <- seurat_objects[[sample_name]]
    

  # Apply QC filters to create a filtered object (adjust thresholds as needed)
    seurat_obj <- subset(seurat_obj, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 10)
    
      ## ---- SAVE VIOLIN PLOT ---- ##
    vln_plot <- VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
    print(vln_plot)
    
    ## ---- SAVE SCATTER PLOT ---- ##
    p1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
    p2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
    scatter_plot <- p1 + p2
    print(scatter_plot)
    
    # Store filtered object in the list
    seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm QC completion
print("QC step completed for all samples.")


```


```{r}

# View metadata for samples

nrow(seurat_objects[["S1_W1_R1"]]@meta.data)
nrow(seurat_objects[["S1_W1_R2"]]@meta.data)
nrow(seurat_objects[["S1_W1_R3"]]@meta.data)
nrow(seurat_objects[["S1_W1_R4"]]@meta.data)

```


########################### Normalizing the data ###############################



```{r normalization_step}

# Loop through each Seurat object for Normalization
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Apply LogNormalization
  seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = 10000)
  
  # Print confirmation for each sample
  print(paste("Normalization completed for:", sample_name))
  
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm completion of normalization for all samples
print("Normalization step completed for all samples.")

```

```{r str_seurat_obj}

str(seurat_obj)

```


############################ Feature selection #################################


```{r feature_selection_step, fig.width=18, fig.height=5}


# Loop through each Seurat object for Feature Selection
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Identify highly variable features (default: 2000 genes, vst method)
  seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)
  
  # Print top 10 highly variable genes
  top10 <- head(VariableFeatures(seurat_obj), 10)
  print(paste("Top 10 HVGs for", sample_name, ":", paste(top10, collapse = ", ")))
  
  # Plot variable features with labels
  plot1 <- VariableFeaturePlot(seurat_obj)
  plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
  print(plot1 + plot2)
  
   
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm feature selection
  print(paste("Feature selection completed for:", sample_name))
}

# Confirm completion of feature selection for all samples
print("Feature selection step completed for all samples.")

```

############################## Scaling the data ################################


```{r scale_data}

# Loop through each Seurat object for Feature Selection
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
 
   # Get all genes
  all.genes <- rownames(seurat_obj)
  
  # Scale the data — scaling all genes
  seurat_obj <- ScaleData(seurat_obj, features = all.genes, verbose = FALSE)
  
  # Update object in the list
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm scaling
  print(paste("Scaling completed for:", sample_name))
}

# Confirm completion
print("Scaling step completed for all samples.")
  
   
```


```{r str_seurat_obj}

str(seurat_obj)

```



##################### linear dimensional reduction (PCA) #######################


```{r PCA_step}

# Loop through each Seurat object for PCA
for (sample_name in names(seurat_objects)) {

  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Run PCA using variable features
  seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(object = seurat_obj))
  print(paste("PCA completed for:", sample_name))
  
  # Print PCA results for first 5 PCs with top 5 features
  print(seurat_obj[["pca"]], dims = 1:5, nfeatures = 5)
  
  # Visualize PCA results using DimHeatmap
  DimHeatmap(seurat_obj, dims = 1, cells = 500, balanced = TRUE)
  
  # Update the object in the list with PCA results
  seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm PCA completion
print("PCA step completed for all samples.")

```



############### Determine the ‘dimensionality’ of the dataset ##################



```{r Determine_DimentionalityofData}

library(ggplot2)

# Loop through each Seurat object to generate and display Elbow Plots
for (sample_name in names(seurat_objects)) {
  
  # Extract the Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Generate Elbow Plot with sample name as title
  elbow_plot <- ElbowPlot(seurat_obj) + ggtitle(paste("Elbow Plot -", sample_name))
  
  # Display the plot
  print(elbow_plot)
}

```



############################# Cluster the cells ################################


```{r Cluster_cells}

# Loop through each Seurat object for clustering
for (sample_name in names(seurat_objects)) {
  
  # Extract the Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Step 1: Find Neighbors (using first 10 PCs)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:7)
  
  # Step 2: Find Clusters (Louvain algorithm with resolution 0.5)
  seurat_obj <- FindClusters(seurat_obj, resolution = 1.0)
  
  # Display a message for progress
  print(paste("Clustering completed for:", sample_name))
  
  # Optional: Display the first few cluster assignments
  print(head(Idents(seurat_obj), 5))
  
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm completion
print("Cell clustering completed for all samples.")


```


############################## UMAP ############################################

```{r non-linear_dimensional_reduction}


# Loop through each Seurat object for UMAP
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Run UMAP using first 10 PCs
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:7)
  
  # Plot UMAP with cluster labels
  umap_plot <- DimPlot(seurat_obj, reduction = "umap", label = TRUE, repel = TRUE) + ggtitle(paste0("UMAP Plot - ", sample_name))
  
  # Display the UMAP plot
  print(umap_plot)
  print(FeaturePlot(seurat_obj, features = "percent.mt"))
  
  # Save UMAP plot as PNG
  plot_file <- paste0(results_dir, "/UMAP_Plots/", sample_name, "_UMAP.png")
  ggsave(plot_file, umap_plot, width = 8, height = 6, dpi = 300)
  print(paste("UMAP plot saved for:", sample_name))
  
  # Save the processed Seurat object as .qs
  qs_file <- paste0(results_dir, "/", sample_name, "_umap.qs")
  qs::qsave(seurat_obj, file = qs_file)
  print(paste("UMAP Seurat object saved for:", sample_name))
  
  # Update object in the list
  seurat_objects[[sample_name]] <- seurat_obj
}

# Confirm completion
print("UMAP and saving completed for all samples.")

```



############################ cluster biomarkers ################################


```{r cluster_biomarker}

# Loop through each Seurat object for UMAP
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Identify markers for cluster 2
  cluster2.markers <- FindMarkers(seurat_obj, ident.1 = 2)

# Print top 5 markers for verification
  print(paste("Top markers for cluster 2 in:", sample_name))
  print(head(cluster2.markers, n = 5))
  
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm completion for each sample
  print(paste("Marker identification completed for:", sample_name))

}

# Confirm completion
print("Cluster marker identification completed for all samples.")


```



####### find all markers distinguishing cluster 5 from clusters 0 and 3 ########


```{r markers_for_ Cluster5}

# Loop through each Seurat object for UMAP
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Identify markers for cluster 2
  cluster5.markers <- FindMarkers(seurat_obj, ident.1 = 5, ident.2 = c(0, 3))

# Print top 5 markers for verification
  print(paste("Top markers for cluster 5 in:", sample_name))
  print(head(cluster5.markers, n = 5))
  
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm completion for each sample
  print(paste("Marker identification completed for:", sample_name))

}

# Confirm completion
print("Cluster marker identification completed for all samples.")


```



###### markers for every cluster compared to all remaining cells (report only positive) ########



```{r findMarkers_allClusters}


# Loop through each Seurat object for UMAP
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  
  #find markers for every cluster compared to all remaining cells, report only the positive ones
  seurat_obj.markers <- FindAllMarkers(seurat_obj, only.pos = TRUE)
  
  # Filter markers with log2 fold change > 1 and store it back
  seurat_obj.markers <- seurat_obj.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
            
  
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm completion for each sample
  print(paste("Marker identification completed for:", sample_name))
  
}


```


##################### tests for differential expression ########################


```{r ROC_test}

# the ROC test returns the ‘classification power’ for any individual marker (ranging from 0 - random, to 1 - perfect).

# Loop through each Seurat object for ROC-based marker identification
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Identify markers for cluster 0 using the ROC test
  cluster0.markers <- FindMarkers(seurat_obj, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)

  # Print top markers for verification
  print(paste("Top ROC-based markers for cluster 0 in:", sample_name))
  print(head(cluster0.markers, n = 5))
  
  # Update the object in the list
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm completion for each sample
  print(paste("ROC-based marker identification completed for:", sample_name))
}

# Confirm completion
print("ROC-based marker identification completed for all samples.")

```





```{r visualize_marker_expression}

# Loop through each Seurat object to visualize marker expression
for (sample_name in names(seurat_objects)) {
  
  # Load Seurat object
  seurat_obj <- seurat_objects[[sample_name]]
  
  # Define marker genes to visualize 
  marker_genes <- c("TPT1", "RPL32")  
  
  # Generate violin plot
  vln_plot <- VlnPlot(seurat_obj, features = marker_genes, ncol = 2) 
  print(vln_plot)
      
  # Generate feature plots on UMAP
  feature_plots <- FeaturePlot(seurat_obj, features = marker_genes) 
  print(feature_plots)
  
  # Update object in the list (no modification, just to maintain consistency)
  seurat_objects[[sample_name]] <- seurat_obj
  
  # Confirm completion for each sample
  print(paste("Marker visualization completed for:", sample_name))
}

```





############################# Annotation-Azimuth ###############################




```{r}

library(Azimuth)

# Loop through each Seurat object for UMAP
for (sample_name in names(seurat_objects)) {
  
  
  
seurat_objects[[sample_name]] <- RunAzimuth(seurat_objects[[sample_name]], reference = "pbmcref")


} 



```


```{r}
# Loop through each Seurat object for UMAP
for (sample_name in names(seurat_objects)) {

predicted.celltype.l1= c("CD4 T", "CD8 T", "other T", "B", "Mono", "NK", "other", "DC")
broad.annotation= c("T", "T", "T", "B", "Mono", "NK", "other", "DC")
 map <- setNames(broad.annotation, predicted.celltype.l1)


seurat_objects[[sample_name]]@meta.data[["broad.annotation"]] <- map[seurat_objects[[sample_name]][["predicted.celltype.l1"]] [, 1]]


}
```



```{r}

seurat_objects[[sample_name]]@meta.data


```

```{r}
?RunAzimuth
```


```{r}

DimPlot(seurat_objects$S1_W1_R1, group.by = "broad.annotation")
DimPlot(seurat_objects$S1_W1_R2, group.by = "broad.annotation")
DimPlot(seurat_objects$S1_W1_R3, group.by = "broad.annotation")
DimPlot(seurat_objects$S1_W1_R4, group.by = "broad.annotation")
```









 













